---
title: "TTE Test: Metformin Initiation in Diabetic Patients (DuckDB Sample Data)"
format: html
editor: visual
---

# Test: Configurable Censoring with Sample Data

This tests the TTE pipeline's configurable censoring feature using the PatientTimelineViewer sample DuckDB data.

**Study Design:** - **Eligibility**: Patients with Type 2 diabetes (E11) and at least 1 prior encounter - **Treatment**: First Metformin prescription - **Outcome**: First heart failure diagnosis (I50) - **Censoring**: Death (testing the new custom censoring feature)

## Setup

```{r}
library(tidyverse)
library(DBI)
library(duckdb)

# Connect to sample DuckDB
db_path <- system.file("extdata", "pcornet_cdw.duckdb", package = "PatientTimelineViewer")
if (db_path == "") {
  db_path <- "/Users/toddjohnson/Documents/GitHub/PatientTimelineViewer/inst/extdata/pcornet_cdw.duckdb"
}
con <- dbConnect(duckdb(), db_path, read_only = TRUE)

# Study parameters
study_start <- "2020-01-01"
study_end <- "2023-12-31"
max_followup_periods <- 12  # months
```

## Step 1: Create Trial Periods (Monthly)

```{r}
# Generate monthly trial periods
start_date <- as.Date(study_start)
end_date <- as.Date(study_end)
n_months <- as.integer(difftime(end_date, start_date, units = "days") / 30.44)

trial_periods <- tibble(
  trial_id = 0:(n_months - 1),
  enrollment_start = seq(start_date, by = "month", length.out = n_months),
  enrollment_end = seq(start_date, by = "month", length.out = n_months) + months(1)
)

# Upload to DuckDB as temp table
dbWriteTable(con, "TrialPeriods", trial_periods, temporary = TRUE, overwrite = TRUE)
print(paste("Created", nrow(trial_periods), "trial periods"))
head(trial_periods)
```

## Step 2: Identify Treatment Events (First Metformin)

```{r}
treatment_sql <- sprintf("
SELECT PATID, MIN(RX_ORDER_DATE) AS treatment_date
FROM PRESCRIBING
WHERE UPPER(RAW_RX_MED_NAME) LIKE '%%METFORMIN%%'
  AND RX_ORDER_DATE <= '%s'
GROUP BY PATID
", study_end)

treatment_events <- dbGetQuery(con, treatment_sql)
dbWriteTable(con, "TreatmentEvents", treatment_events, temporary = TRUE, overwrite = TRUE)

print(paste("Patients with Metformin:", nrow(treatment_events)))
print(treatment_events)
```

## Step 3: Create Patient Events (Encounters during study period)

```{r}
events_sql <- sprintf("
SELECT
    D.PATID AS UID,
    D.PATID,
    E.ENCOUNTERID,
    E.ADMIT_DATE AS event_date,
    'ENCOUNTER' AS event_type,
    E.ENC_TYPE
FROM DEMOGRAPHIC AS D
INNER JOIN ENCOUNTER AS E ON E.PATID = D.PATID
WHERE E.ADMIT_DATE BETWEEN '%s' AND '%s'
", study_start, study_end)

patient_events <- dbGetQuery(con, events_sql)
dbWriteTable(con, "PatientEvents", patient_events, temporary = TRUE, overwrite = TRUE)

print(paste("Total events:", nrow(patient_events)))
print(paste("Unique patients with events:", n_distinct(patient_events$PATID)))
```

## Step 4: Add Treatment Status and Prior Encounter Count

```{r}
events_with_status_sql <- "
SELECT
    PE.*,
    (SELECT COUNT(*) FROM PatientEvents PE2
     WHERE PE2.UID = PE.UID AND PE2.event_date < PE.event_date) AS prior_encounter_count,
    TE.treatment_date AS FirstTreatmentDT,
    CASE
        WHEN TE.treatment_date IS NOT NULL AND TE.treatment_date < PE.event_date
        THEN 1 ELSE 0
    END AS has_prior_treatment
FROM PatientEvents AS PE
LEFT JOIN TreatmentEvents AS TE ON TE.PATID = PE.PATID
"

events_with_status <- dbGetQuery(con, events_with_status_sql)
dbWriteTable(con, "EventsWithStatus", events_with_status, temporary = TRUE, overwrite = TRUE)

print("Treatment status distribution:")
events_with_status %>%
  group_by(has_prior_treatment) %>%
  summarise(n_events = n(), n_patients = n_distinct(UID))
```

## Step 5: Filter to Eligible Time Zeros (Diabetic patients with prior encounters)

```{r}
# Eligibility: Has diabetes diagnosis AND at least 1 prior encounter AND no prior treatment
eligible_sql <- sprintf("
SELECT EWS.*,
    CASE WHEN EWS.FirstTreatmentDT = EWS.event_date THEN 1 ELSE 0 END AS treatment_today
FROM EventsWithStatus AS EWS
WHERE EWS.event_date >= '%s'
  AND EWS.prior_encounter_count >= 1
  AND EWS.has_prior_treatment = 0
  AND EXISTS (
    SELECT 1 FROM DIAGNOSIS AS DX
    WHERE DX.PATID = EWS.PATID
      AND DX.DX LIKE 'E11%%'
      AND DX.DX_DATE < EWS.event_date
  )
", study_start)

eligible_time_zeros <- dbGetQuery(con, eligible_sql)
dbWriteTable(con, "EligibleTimeZeros", eligible_time_zeros, temporary = TRUE, overwrite = TRUE)

print(paste("Eligible time zeros:", nrow(eligible_time_zeros)))
print("By treatment status:")
eligible_time_zeros %>%
  group_by(treatment_today) %>%
  summarise(n = n(), n_patients = n_distinct(UID))
```

## Step 6: Assign to Trial Periods

```{r}
assignment_sql <- "
SELECT
    ETZ.UID, ETZ.PATID, ETZ.event_date AS time_zero,
    ETZ.treatment_today AS treatment,
    TP.trial_id, TP.enrollment_start, TP.enrollment_end,
    ETZ.FirstTreatmentDT
FROM EligibleTimeZeros AS ETZ
INNER JOIN TrialPeriods AS TP
    ON ETZ.event_date >= TP.enrollment_start
    AND ETZ.event_date < TP.enrollment_end
"

trial_assignments <- dbGetQuery(con, assignment_sql)
dbWriteTable(con, "TrialAssignments", trial_assignments, temporary = TRUE, overwrite = TRUE)

print(paste("Trial assignments:", nrow(trial_assignments)))
trial_assignments %>%
  group_by(treatment) %>%
  summarise(n = n(), n_patients = n_distinct(UID))
```

## Step 7: Filter to One Assignment per Patient-Trial (keep treatment if both)

```{r}
filtered_sql <- "
WITH RankedAssignments AS (
    SELECT *,
        ROW_NUMBER() OVER (
            PARTITION BY UID, trial_id
            ORDER BY treatment DESC, time_zero
        ) AS row_priority
    FROM TrialAssignments
)
SELECT UID, PATID, time_zero, treatment, trial_id,
       enrollment_start, enrollment_end, FirstTreatmentDT
FROM RankedAssignments
WHERE row_priority = 1
"

filtered_assignments <- dbGetQuery(con, filtered_sql)

# Remove trials after first treatment
final_sql <- "
WITH FirstTreatmentTrial AS (
    SELECT UID, MIN(CASE WHEN treatment = 1 THEN trial_id END) AS first_treatment_trial_id
    FROM (%s) AS FA
    GROUP BY UID
)
SELECT FA.*
FROM (%s) AS FA
LEFT JOIN FirstTreatmentTrial AS FTT ON FTT.UID = FA.UID
WHERE FTT.first_treatment_trial_id IS NULL
   OR FA.trial_id <= FTT.first_treatment_trial_id
"

final_assignments <- dbGetQuery(con, sprintf(final_sql, filtered_sql, filtered_sql))
dbWriteTable(con, "FinalTrialAssignments", final_assignments, temporary = TRUE, overwrite = TRUE)

print(paste("Final assignments:", nrow(final_assignments)))
final_assignments %>%
  group_by(treatment) %>%
  summarise(n = n(), n_patients = n_distinct(UID))
```

## Step 8: Identify Outcomes (Heart Failure - I50)

```{r}
outcome_sql <- sprintf("
SELECT FTA.UID, FTA.trial_id, FTA.time_zero, MIN(DX.DX_DATE) AS outcome_date
FROM FinalTrialAssignments AS FTA
INNER JOIN DIAGNOSIS AS DX ON DX.PATID = FTA.PATID
WHERE DX.DX_DATE > FTA.time_zero
  AND DX.DX_DATE <= '%s'
  AND DX.DX LIKE 'I50%%'
GROUP BY FTA.UID, FTA.trial_id, FTA.time_zero
", study_end)

outcomes <- dbGetQuery(con, outcome_sql)
dbWriteTable(con, "Outcomes", outcomes, temporary = TRUE, overwrite = TRUE)

print(paste("Patients with outcome:", nrow(outcomes)))
if (nrow(outcomes) > 0) print(outcomes)
```

## Step 9: Last Visit for Censoring

```{r}
last_visit_sql <- "
SELECT PATID AS UID, MAX(ADMIT_DATE) AS last_visit_date
FROM ENCOUNTER
GROUP BY PATID
"

last_visit <- dbGetQuery(con, last_visit_sql)
dbWriteTable(con, "LastVisit", last_visit, temporary = TRUE, overwrite = TRUE)
print(paste("Patients with last visit:", nrow(last_visit)))
```

## Step 10b: Custom Censoring Events (DEATH - Testing New Feature!)

**This is the key test of the configurable censoring feature**

```{r}
# Custom censoring query - deaths within study period
censoring_sql <- sprintf("
SELECT PATID AS UID, DEATH_DATE AS censor_date, 'death' AS censor_reason
FROM DEATH
WHERE DEATH_DATE <= '%s'
  AND DEATH_DATE IS NOT NULL
", study_end)

censoring_events <- dbGetQuery(con, censoring_sql)
dbWriteTable(con, "CensoringEvents", censoring_events, temporary = TRUE, overwrite = TRUE)

print("=== CUSTOM CENSORING EVENTS (DEATH) ===")
print(paste("Patients with death censoring:", nrow(censoring_events)))
if (nrow(censoring_events) > 0) {
  print(censoring_events)
}
```

## Step 11: Create Base Dataset with Custom Censoring

```{r}
base_dataset_sql <- sprintf("
SELECT
    FTA.UID, FTA.PATID, FTA.trial_id,
    FTA.enrollment_start AS trial_start_date,
    FTA.enrollment_end AS trial_end_date,
    FTA.time_zero,
    FTA.treatment,
    FTA.FirstTreatmentDT,

    -- Outcome date
    CASE
        WHEN O.outcome_date IS NOT NULL AND O.outcome_date > FTA.time_zero
        THEN O.outcome_date ELSE NULL
    END AS outcome_date,

    -- Custom censoring date (DEATH)
    CASE
        WHEN CE.censor_date IS NOT NULL AND CE.censor_date > FTA.time_zero
        THEN CE.censor_date ELSE NULL
    END AS custom_censor_date,

    -- Outcome indicator
    CASE
        WHEN O.outcome_date IS NOT NULL AND O.outcome_date > FTA.time_zero
        THEN 1 ELSE 0
    END AS outcome,

    -- Censoring reason (priority cascade)
    CASE
        WHEN O.outcome_date IS NOT NULL AND O.outcome_date > FTA.time_zero
        THEN 'outcome'
        WHEN CE.censor_date IS NOT NULL AND CE.censor_date > FTA.time_zero
        THEN CE.censor_reason  -- 'death' from custom censoring
        WHEN LV.last_visit_date < '%s'
        THEN 'last_visit'
        ELSE 'administrative'
    END AS censor_reason,

    LV.last_visit_date

FROM FinalTrialAssignments AS FTA
LEFT JOIN Outcomes AS O
    ON O.UID = FTA.UID AND O.trial_id = FTA.trial_id
LEFT JOIN LastVisit AS LV
    ON LV.UID = FTA.UID
LEFT JOIN CensoringEvents AS CE
    ON CE.UID = FTA.UID
", study_end)

base_dataset <- dbGetQuery(con, base_dataset_sql)

print("=== BASE DATASET WITH CUSTOM CENSORING ===")
print(paste("Total rows:", nrow(base_dataset)))

print("\n=== CENSORING REASON BREAKDOWN ===")
censor_breakdown <- base_dataset %>%
  group_by(censor_reason) %>%
  summarise(
    n_observations = n(),
    n_patients = n_distinct(UID)
  ) %>%
  arrange(desc(n_observations))
print(censor_breakdown)

# Show death-censored patients specifically
death_censored <- base_dataset %>% filter(censor_reason == "death")
if (nrow(death_censored) > 0) {
  print("\n=== DEATH-CENSORED OBSERVATIONS (Custom Censoring Working!) ===")
  print(death_censored %>% select(UID, trial_id, time_zero, custom_censor_date, censor_reason))
}
```

## Summary: Censoring Feature Test Results

```{r}
cat("\n========================================\n")
cat("CENSORING FEATURE TEST SUMMARY\n")
cat("========================================\n\n")

cat("Study Design:\n")
cat("- Treatment: Metformin initiation\n")
cat("- Outcome: Heart failure (I50)\n")
cat("- Custom Censoring: Death\n\n")

cat("Results:\n")
cat(sprintf("- Total trial assignments: %d\n", nrow(base_dataset)))
cat(sprintf("- Unique patients: %d\n", n_distinct(base_dataset$UID)))
cat(sprintf("- Deaths in sample data: %d\n", nrow(censoring_events)))
cat(sprintf("- Death-censored observations: %d\n", sum(base_dataset$censor_reason == "death")))

if (sum(base_dataset$censor_reason == "death") > 0) {
  cat("\n✓ CUSTOM CENSORING FEATURE IS WORKING!\n")
  cat("  Death records are correctly being used to censor patients.\n")
} else if (nrow(censoring_events) == 0) {
  cat("\n⚠ No deaths in the eligible patient subset.\n")
  cat("  (This is expected if deceased patients weren't in the diabetic cohort)\n")
} else {
  cat("\n⚠ Deaths exist but no observations were censored.\n")
  cat("  (Deaths may have occurred before patient's time_zero)\n")
}
```

## Cleanup

```{r}
dbDisconnect(con)
cat("\nDatabase connection closed.\n")
```
